---
title: "Untitled"
author: "Mike McPhee Anderson, FSA"
date: "2025-01-01"
output: html_document
---

## Mortality Improvement

### Non-smokers

#### Evaluate knot locations

Very heavily skewed towards old age deaths, looks like first knot should be
places around 50 (5% of the deaths), second around 65.

```{r}

plot_data <- df_study_data %>%
  filter(
    smoker_status == "NONSMOKER",
    attained_age >= 25,
    attained_age <= 90) %>%
  group_by(attained_age, gender) %>%
  summarise(
    n_deaths = sum(number_of_deaths)
  ) %>%
  ungroup() %>%
  group_by(gender) %>%
  arrange(attained_age) %>%
  mutate(
    cumulative_pct_of_deaths = cumsum(n_deaths) / sum(n_deaths)
  ) 

ggplot(plot_data, aes(x=attained_age, y=cumulative_pct_of_deaths, group=gender, color=gender)) +
  geom_line() +
  scale_y_continuous(
    labels = scales::percent_format(), 
    transform = scales::pseudo_log_trans(sigma = 0.01),
    breaks=seq(0,1,by=0.1)) +
  scale_x_continuous(breaks=seq(0, 120, by=10)) +
  # eye-ball first knot location
  geom_hline(yintercept = 0.05) +
  geom_vline(xintercept = 50) +
  # eye-ball second knot location
  geom_hline(yintercept = 0.15) +
  geom_vline(xintercept = 65) +
  xlab("Attained Age") +
  ylab("Cumulative % of Deaths") +
  theme_bw()

```

### Build inferential model

By accounting for the other time variables (attained age + duration), we can see
what is left over by observation (calendar) year.  This will give us a starting 
point for a mortality improvement assumption.  Cap the duration at 20, since 
underwriting should largely be worn-off by that time.

```{r}

df_model_data <- df_study_data %>%
  filter(smoker_status == "NONSMOKER",
         number_of_preferred_classes != "NA",
         attained_age >= 25,
         attained_age <= 90) %>%
  mutate(
    gender = factor(gender, levels=c("MALE", "FEMALE")),
    capped_duration = pmin(20, duration),
    expected_claims = qx * policies_exposed,
    face_amount_band = ordered(face_amount_band, levels=!!face_amount_order),
    face_amount_band_int = as.integer(face_amount_band)
  ) %>%
  filter(face_amount_band_int > 2) %>%
  group_by(observation_year, gender, attained_age, capped_duration, face_amount_band, face_amount_band_int) %>%
  summarise(
    number_of_deaths = sum(number_of_deaths),
    expected_claims = sum(expected_claims),
    .groups="drop") %>%
  ungroup() %>%
  filter(expected_claims > 0)

X_mat <- model.matrix(
  number_of_deaths ~ gender * 
    splines::ns(attained_age, Boundary.knots = c(25, 90), knots = c(35, 45, 55, 65, 75, 85)) *
    splines::ns(capped_duration, Boundary.knots=c(1, 20), knots = c(5, 10, 15)) - 1,
  data=df_model_data
)

X_offset <- log(df_model_data$expected_claims)

y_mat <- df_model_data$number_of_deaths

```

#### Run CV

Select the lambda for the lasso based on a rudimentary CV.

```{r}

cv_model <- cv.glmnet(
  X_mat,
  y_mat,
  family="poisson",
  offset=X_offset,
  fit_intercept=T
)

cv_model

```

#### Build Model

Fit the model and plot the lambda selected via CV

```{r}

glmnet_model <- glmnet::glmnet(
  X_mat,
  y_mat,
  family="poisson",
  offset=X_offset,
  fit_intercept=T
)

lambda_diff <- abs(glmnet_model$lambda - cv_model$lambda.1se)

LAMBDA_INDEX <- which(lambda_diff == min(lambda_diff))

best_lambda_value <- glmnet_model$lambda[LAMBDA_INDEX]

plot(glmnet_model$dev.ratio, type="l")
points(LAMBDA_INDEX, glmnet_model$dev.ratio[LAMBDA_INDEX])

```


#### Check Model

```{r}

model_pred <- predict(
  glmnet_model,
  newx=X_mat,
  newoffset = X_offset,
  s=best_lambda_value,
  type = "response")

df_model_data$model_pred <- model_pred

# sanity check
sum(df_model_data$number_of_deaths) / sum(model_pred)

```

#### Look for any major areas of poor fit

Use a decision tree to evaluate the model fit / residuals on the training set.

* Early durations don't look great, but there aren't a lot of deaths

```{r}

model_errors <- as.matrix(df_model_data[,c("model_pred","number_of_deaths")])

rpart_model <- rpart(
  model_errors ~ capped_duration + attained_age + gender,
  data=df_model_data,
  method="poisson",
  control = rpart.control(cp=0.001, maxdepth = 3))

rpart.plot(rpart_model, digits=3)

```

#### Check Mortality Improvement

Downward sloping lines mean that mortality is getting better over time, this
is called *mortality improvement*.  It's generally attributed to changes in lifestyle (fewer smokers),
and medical advances (statins, better cancer treatment, etc.).

* Evidence of improvement between 55 and 75

* The upward sloping lines at earlier ages is consistent with *deaths of despair* that occured following the 2008 recession, where mortality worsened for younger ages.


```{r}

age_cuts <- c(-Inf, seq(35, 85, by=10), Inf)

plot_data <- df_model_data %>%
  filter(observation_year < 2018) %>%
  mutate(
    age_group = cut(attained_age, breaks=!!age_cuts)
  ) %>%
  group_by(age_group, gender, observation_year) %>%
  summarise(
    number_of_deaths = sum(number_of_deaths),
    model_pred = sum(model_pred),
    .groups="drop"
  ) %>%
  ungroup()

ggplot(plot_data, aes(x=observation_year, y=log(number_of_deaths/model_pred), color=gender, group=gender)) +
  facet_wrap(~ age_group, scales="free_y") +
  geom_line() +
  geom_smooth(method="lm", se=F) +
  ggtitle("Mortality Improvement Check") +
  theme_bw() +
  xlab("Observation Year") +
  ylab("Log A/E")

```

No improvement in 75-85, regardless of the face amounts

```{r}

plot_data <- df_model_data %>%
  filter(observation_year < 2018) %>%
  mutate(
    age_group = cut(attained_age, breaks=!!age_cuts)
  ) %>%
  filter(age_group == "(75,85]") %>%
  group_by(face_amount_band, gender, observation_year) %>%
  summarise(
    number_of_deaths = sum(number_of_deaths),
    model_pred = sum(model_pred),
    .groups="drop"
  ) %>%
  ungroup()

ggplot(plot_data, aes(x=observation_year, y=log(number_of_deaths/model_pred), color=gender, group=gender)) +
  facet_wrap(~ face_amount_band, scales="free_y") +
  geom_line() +
  geom_smooth(method="lm", se=F) +
  ggtitle("Mortality Improvement Check - 75 to 84") +
  theme_bw() +
  xlab("Observation Year") +
  ylab("Log A/E")

```

No strong evidence of improvement at 85+

```{r}

plot_data <- df_model_data %>%
  filter(observation_year < 2018) %>%
  mutate(
    age_group = cut(attained_age, breaks=!!age_cuts)
  ) %>%
  filter(age_group == "(85, Inf]") %>%
  group_by(face_amount_band, gender, observation_year) %>%
  summarise(
    number_of_deaths = sum(number_of_deaths),
    model_pred = sum(model_pred),
    .groups="drop"
  ) %>%
  ungroup()

ggplot(plot_data, aes(x=observation_year, y=log(number_of_deaths/model_pred), color=gender, group=gender)) +
  facet_wrap(~ face_amount_band, scales="free_y") +
  geom_line() +
  geom_smooth(method="lm", se=F) +
  ggtitle("Mortality Improvement Check - 85+") +
  theme_bw() +
  xlab("Observation Year") +
  ylab("Log A/E")

```



