---
title: "Perm Mortality Model"
author: "Mike McPhee Anderson, FSA"
date: "2024-12-18"
output: html_document
---

```{r setup, include=FALSE}

library(glmnet)
library(xgboost)
library(recipes)
library(rpart)
library(rpart.plot)
library(splines)
library(rsample)
library(shapviz)
library(tidyselect)
library(doMC)
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)

```

# Overview

## Models

Within the ILEC data "Perm" (permanent) life insurance category, there
are some major differences in the underwriting / target market that can be inferred via EDA. 
These differences lead us to three distinct models:

 1. *Simplified Issue:* Policies issued after 2005 without any preferred classes are likely simplified
   issue.  These policies are less than $100K face amount, and are likely to have less stringent underwriting.
   
 2. *Fully Underwritten (recent):*  Post-2005 policies are likely to have at least one preferred class, and I recall that blood testing gained traction in the late 1990s.  These policies will also have some underwriting class selection (select period) that is still applicable.
 
 3. *Fully Underwritten (historical):*  Policies prior to 2005 are likely to have little remaining effects from underwriting selection.  Also, this data is unlikely to have preferred classes, and is mostly just smoker vs. non-smoker for underwriting classes.  The majority of it will be smaller face amounts.
 
## Training and Testing Splits

Unfortunately, the ILEC data is aggregated in such a way that makes a typical train / test split cumbersome. While we could sample the exposures and deaths for each aggregated cell, things like fractional exposures and the actuarial custom of setting deaths to have a full year of exposure introduce some opportunities for noise.  

Instead, we'll use the most recent two years of data for testing, and perform cross-validation within the training set by observation year.  This is more closely aligned with what an "unseen observation" is in practice, i.e., another calendar year of data.  

# Setup 

## Load Data

```{r}

face_amount_order <- c(
  "1-9999",
  "10000-24999",
  "25000-49999",
  "50000-99999",
  "100000-249999",
  "250000-499999",
  "500000-999999",
  "1000000-2499999",
  "2500000-4999999",
  "5000000-9999999",
  "10000000+"
)

df_study_data <- read_rds("ilec_perm_recent.rds") %>%
  mutate(
    train_test = ifelse(observation_year >= 2016, "TEST", "TRAIN"),
      # no selection period, just ultimate mortality
    expected_deaths_08_vbt_ult = coalesce(qx * policies_exposed),
    # create ordered factor + integer representation of face amount bands
    face_amount_band = ordered(face_amount_band, levels=!!face_amount_order),
    face_amount_band_int = as.integer(face_amount_band),
    issue_age = attained_age - duration + 1
  ) %>%
  filter(expected_deaths_08_vbt_ult > 0)

```

# Perm Historical Model

## Train Test Split

```{r}

df_ph_data <- df_study_data 

df_ph_data.train <- df_ph_data %>% 
  filter(train_test == "TRAIN")

df_ph_data.test <- df_ph_data %>%
  filter(train_test == "TEST")

df_ph_data %>%
  group_by(train_test) %>%
  summarise(
    n_deaths = sum(number_of_deaths),
    .groups="drop"
  ) %>%
  ungroup() %>%
  mutate(
    pct_deaths = n_deaths / sum(n_deaths)
  )

```

## Xgboost Model

```{r}

df_ph_data.train %>%
  group_by(smoker_status, number_of_preferred_classes, preferred_class) %>%
  summarise(
    n_deaths = sum(number_of_deaths)
  )

```

```{r}

summary(df_ph_data.train)

```

### Knot Selection

```{r}

group_model_data <- function(df) {
  df %>%
    group_by(attained_age, duration, issue_age, number_of_preferred_classes,
             preferred_class, smoker_status, face_amount_band_int, gender) %>%
    summarise(
      number_of_deaths = sum(number_of_deaths),
      expected_deaths_08_vbt_ult = sum(expected_deaths_08_vbt_ult),
      .groups="drop"
    ) %>%
    ungroup()
}

model_data.train <- df_ph_data.train %>% group_model_data()
model_data.test <- df_ph_data.test %>% group_model_data()

# ph.glmnet.recipe <- recipe(number_of_deaths ~ face_amount_band_int + 
#                              attained_age + 
#                              duration + 
#                              issue_age +
#                              gender +
#                              number_of_preferred_classes +
#                              preferred_class +
#                              smoker_status,
#                            data = model_data.train) %>%
#   step_my_model_factor_prep() %>%
#   prep(model_data.train, retain=F)
# 
# bake(ph.glmnet.recipe, model_data.train) %>%
#   summary()

```


```{r}

# source("0_lib.R")
# 
# knot_loc_ptiles <- seq(0.05, 0.95, by=0.1)
# 
# attage_ecdf <- model_data.train %>%
#   approx_ecdf(attained_age, number_of_deaths)
# 
# attage_knots <- unique(attage_ecdf(knot_loc_ptiles))
# 
# face_amt_ecdf <- model_data.train %>%
#   approx_ecdf(face_amount_band_int, number_of_deaths)
# 
# face_amt_knots <- unique(face_amt_ecdf(knot_loc_ptiles))
# 
# duration_knots <- log(c(10, 15))
# 
# issue_age_ecdf <- model_data.train %>%
#   approx_ecdf(issue_age, number_of_deaths)
# 
# issue_age_knots <- c(28, 38, 45, 50, 54, 58, 62, 66, 70)
# 
# ATTAGE_PART = "splines::ns(attained_age, Boundary.knots = c(25, 95), knots=c(%KNOTS%))"
# FACE_AMT_PART = "splines::bs(face_amount_band_int, degree=1, Boundary.knots = c(1, 11), knots=c(%KNOTS%))"
# DUR_PART = "splines::bs(log(duration), degree=1, Boundary.knots = c(0, log(18)), knots=c(%KNOTS%))"
# ISSAGE_PART = "splines::ns(issue_age, Boundary.knots = c(18, 75), knots=c(%KNOTS%))"
# 
# formula_template <- paste0(
#   "number_of_deaths ~ preferred_class + smoker_status + ",
#   stringr::str_glue("gender*{ATTAGE_PART} + {FACE_AMT_PART} + {DUR_PART}*{ISSAGE_PART}"),
#   " -1"
# )
# 
# spline_locations <- list()
# 
# spline_locations[[ATTAGE_PART]] <- attage_knots
# spline_locations[[FACE_AMT_PART]] <- face_amt_knots
# spline_locations[[DUR_PART]] <- duration_knots 
# spline_locations[[ISSAGE_PART]] <- issue_age_knots 

```

```{r}
# 
# N_BITS <- length(attage_knots) + 
#   length(face_amt_knots) + 
#   length(duration_knots) +
#   length(issue_age_knots)
# 
# bit_str_to_formula <- make_bit_str_to_formula(formula_template, spline_locations)
# 
# ga_fitness <- function(bit_string) {
#   
#   this_formula <- bit_str_to_formula(bit_string)
#   
#   X_mat <- model.matrix(this_formula, data=model_data.train)
#   
#   df_wts <- tibble(
#     term_name = colnames(X_mat)) %>%
#     mutate(
#       lower_limits = case_when(
#         stringr::str_detect(term_name, fixed("log(duration)")) ~ 0,
#         TRUE ~ -Inf
#       ),
#       upper_limits = case_when(
#         (stringr::str_detect(term_name, fixed("issue_age"))) &
#           !(stringr::str_detect(term_name, fixed("log(duration)"))) ~ 0,
#         TRUE ~ Inf
#       )
#     )
# 
#   glmnet_fit <- glmnet::glmnet(
#     X_mat,
#     model_data.train$number_of_deaths,
#     offset = log(model_data.train$expected_deaths_08_vbt_ult),
#     family = "poisson",
#     intercept = T,
#     lower.limits = df_wts$lower_limits,
#     upper.limits = df_wts$upper_limits
#   )
#   
#   aic_bic <- calc_aic_bic_glmnet(glmnet_fit)
#   
#   -1 * min(aic_bic$BIC)
# }


```

```{r}

# memo_fn_fitness <- memoise::memoise(ga_fitness)
# 
# best_model <- GA::ga(type="binary", 
#        memo_fn_fitness, 
#        popSize=20, 
#        nBits = N_BITS, 
#        maxiter = 5, 
#        parallel = F,
#        pmutation = 0.8,
#        elitism = 1,
#        suggestions = rbind(
#          rep(0, N_BITS),
#          rep(1, N_BITS)
#        ),
#        pcrossover=0.7)
# 
# bit_str_to_formula(best_model@solution)

```


```{r}

# starting_sols <-matrix(
#   rep(best_model@solution, 10), 
#   nrow=10, 
#   ncol=length(best_model@solution), byrow = T)
# 
# best_model <- GA::ga(type="binary", 
#        memo_fn_fitness, 
#        popSize=10, 
#        nBits = N_BITS, 
#        maxiter = 10, 
#        parallel = F,
#        pmutation = 0.3,
#        elitism = 2,
#        suggestions = starting_sols,
#        pcrossover=0.0)
# 
# bit_str_to_formula(best_model@solution)

```


### Setup GLMNet recipe

```{r}

# DURATION_INTERP_START <- 10
# DURATION_INTERP_END <- 15
# DURATION_INTERP_POWER <- 0.33
# 
# prep_log_interp <- function(df) {
#   df %>%
#     mutate(
#       c1 = pmin(1, pmax(0, DURATION_INTERP_END - duration) / (DURATION_INTERP_END - DURATION_INTERP_START))^DURATION_INTERP_POWER,
#       c2 = 1 - c1,
#       log_duration_interp = c1 * log(duration) + c2 * log(DURATION_INTERP_END)
#     ) %>%
#     select(-c1, -c2)
# }
# 
# model_data.train <- model_data.train %>% prep_log_interp()
# model_data.test <-  model_data.test %>% prep_log_interp()
# 
# best_formula <- number_of_deaths ~ 
#     preferred_class + 
#     smoker_status + 
#     gender * 
#       splines::ns(attained_age, Boundary.knots = c(25, 95), knots = c(35, 45, 52, 70)) +
#     splines::bs(face_amount_band_int, degree = 1, Boundary.knots = c(1, 11), knots = c(2, 7)) + 
#     log_duration_interp * splines::ns(issue_age, Boundary.knots = c(18, 75), knots = c(45, 58, 66, 70)) - 1
# 
# X_mat <- model.matrix(best_formula, data=model_data.train)
# 
# df_wts <- tibble(
#   term_name = colnames(X_mat)) %>%
#   mutate(
#     lower_limits = case_when(
#       stringr::str_detect(term_name, fixed("log(duration)")) ~ 0,
#       TRUE ~ -Inf
#     ),
#     upper_limits = case_when(
#       (stringr::str_detect(term_name, fixed("issue_age"))) &
#         !(stringr::str_detect(term_name, fixed("log(duration)"))) ~ 0,
#       TRUE ~ Inf
#     )
#   )
# 
# final_model <- glmnet::glmnet(
#     X_mat,
#     model_data.train$number_of_deaths,
#     offset = log(model_data.train$expected_deaths_08_vbt_ult),
#     family = "poisson",
#     intercept = T,
#     lower.limits = df_wts$lower_limits,
#     upper.limits = df_wts$upper_limits
#   )
#   
# aic_bic <- calc_aic_bic_glmnet(final_model)
# 
# min(aic_bic$BIC)

```


### Plot Deviance Ratio vs. Lambda

```{r}

# BEST_LAMBDA_IDX <- 35
# 
# best_lambda_val <- final_model$lambda[BEST_LAMBDA_IDX]
# plot(final_model$dev.ratio, type="l")
# points(BEST_LAMBDA_IDX, final_model$dev.ratio[BEST_LAMBDA_IDX])

```


### Examine model factors

```{r}
# 
# best_coefs <- predict.glmnet(
#   best_fit, newx=X_mat, s=best_lambda_val, type = "coefficients"
# )
# 
# predictor_groups <- tibble(
#   term_name = best_coefs@Dimnames[[1]]
#   ) %>%
#   mutate(
#     term_group = case_when(
#       stringr::str_detect(term_name, fixed("Intercept")) ~ "factor_base_rate",
#       stringr::str_detect(term_name, fixed("preferred_class")) ~ "factor_preferred_class",
#       stringr::str_detect(term_name, fixed("smoker_status")) ~ "factor_smoker_status",
#       stringr::str_detect(term_name, fixed("gender")) |
#         stringr::str_detect(term_name, fixed("attained_age")) ~ "factor_gender_x_attained_age",
#       stringr::str_detect(term_name, fixed("issue_age")) |
#         stringr::str_detect(term_name, fixed("duration")) ~ "factor_issue_age_x_duration",
#       stringr::str_detect(term_name, fixed("face_amount_band_int")) ~ "factor_face_amount"
#     )
#   )
# 
# predictor_groups %>% view()
```



```{r}

# pred_grid <- expand.grid(
#   face_amount_band_int = seq(min(df_ph_data.train$face_amount_band_int), max(df_ph_data.train$face_amount_band_int), by=1),
#   attained_age = seq(min(df_ph_data.train$attained_age), max(df_ph_data.train$attained_age), by=1),
#   duration = seq(min(df_ph_data.train$duration), max(df_ph_data.train$duration), by=1),
#   issue_age = seq(min(df_ph_data.train$issue_age), max(df_ph_data.train$issue_age), by=1),
#   gender = unique(df_ph_data.train$gender),
#   number_of_preferred_classes = unique(df_ph_data.train$number_of_preferred_classes),
#   preferred_class = unique(df_ph_data.train$preferred_class),
#   smoker_status = unique(df_ph_data.train$smoker_status),
#   number_of_deaths = 0, 
#   stringsAsFactors = F) %>% 
#   filter(
#     issue_age + duration - 1 == attained_age,
#     ((smoker_status == "NONSMOKER") | ((smoker_status == "SMOKER") & (number_of_preferred_classes == "2"))),
#     (
#       ((number_of_preferred_classes == "2") & (preferred_class %in% c("1", "2"))) |
#       ((number_of_preferred_classes == "3") & (preferred_class %in% c("1", "2", "3")))
#     )) %>%
#   prep_log_interp()
# 
# X_pred_grid <- cbind(
#   rep(1, nrow(pred_grid)),
#   model.matrix(best_formula, data=pred_grid)
# )
# 
# coeff_diag <- rep(0, best_coefs@Dim[1])
# coeff_diag[best_coefs@i + 1] <- best_coefs@x
# coeff_diag <- diag(coeff_diag)
# 
# pred_mat <- X_pred_grid %*% coeff_diag
# colnames(pred_mat) <- best_coefs@Dimnames[[1]]
# 
# pred_df <- pred_mat %>%
#   as.data.frame() %>%
#   mutate(
#     pred_idx = row_number()) %>%
#   pivot_longer(
#     cols=-c(pred_idx), 
#     names_to = "term_name", 
#     values_to = "term_value") %>%
#   inner_join(predictor_groups, by="term_name") %>%
#   group_by(pred_idx, term_group) %>%
#   summarise(
#     total_value = exp(sum(term_value)),
#     .groups = "drop"
#   ) %>%
#   ungroup() %>%
#   pivot_wider(
#     id_cols = pred_idx, names_from = "term_group", values_from = "total_value"
#   )
# 
# full_pred_df <- bind_cols(
#   pred_grid,
#   pred_df
# )

```

```{r}

# plot_data <- full_pred_df %>%
#   select(issue_age, duration, factor_issue_age_x_duration, factor_base_rate) %>%
#   distinct()
# 
# ggplot(plot_data, aes(x=duration, y=factor_issue_age_x_duration, group=issue_age, color=issue_age)) +
#   geom_line()

```

### Evaluate Model

```{r}

# colnames(model_data.train)

```


```{r}

# BEST_LAMBDA_IDX <- 35
# 
# best_lambda_val <- ga_model$lambda[BEST_LAMBDA_IDX]
# plot(ga_model$dev.ratio, type="l")
# points(BEST_LAMBDA_IDX, ga_model$dev.ratio[BEST_LAMBDA_IDX])
# 
# aic_bic$BIC

```


### Check Model w/ Decision Tree

```{r fig.width=10}

# prep_rpart_data <- function(df) {
#   X <- df %>% 
#     select(-number_of_deaths, 
#            -expected_deaths_08_vbt_ult, 
#            -log_duration_interp,
#            -pred_deaths)
#   
#   y <- df[, c("pred_deaths", "number_of_deaths")] %>%
#     as.matrix()
#   
#   list(
#     "X" = X,
#     "y" = y
#   )
# }
# 
# rpart_data <- model_data.train %>%
#   prep_rpart_data
# 
# model_errors <- rpart_data$y
# 
# rpart_model <- rpart(
#   model_errors ~ .,
#   data=rpart_data$X,
#   method="poisson",
#   control = rpart.control(cp=0.0001, maxdepth = 3))
# 
# rpart.plot(rpart_model, digits=4)

```

```{r fig.width=10}

# rpart_data <- model_data.test %>%
#   prep_rpart_data()
# 
# model_errors <- rpart_data$y
# 
# rpart_model <- rpart(
#   model_errors ~ .,
#   data=rpart_data$X,
#   method="poisson",
#   control = rpart.control(cp=0.0001, maxdepth = 3))
# 
# rpart.plot(rpart_model, digits=3)


```



```{r}

interp_spline <- function(v, interp_start, interp_end, interp_power) {
  c1 <- pmin(1, pmax(0, interp_end - v) / (interp_end - interp_start))^interp_power
  c1 * log(v) + (1-c1) * log(interp_end)
}

issue_age_groups <- list(
  c(-Inf, seq(35, 75, by=10) ,Inf),
  c(-Inf, 45, 55, 65, 75, Inf),
  c(-Inf, 25, 35, 45, 55, 65, 75, Inf),
  c(-Inf, 45, 55, 65, 70, Inf)
)

N_ISSUE_AGE_BITS <- log2(length(issue_age_groups))

interp_params <- list(
  
  c(5, 15, 0.5),
  c(7, 15, 0.5),
  c(10, 15, 0.5),
  c(13, 15, 0.5),
  
  c(5, 15, 0.66),
  c(7, 15, 0.66),
  c(10, 15, 0.66),
  c(13, 15, 0.66),
  
  c(5, 10, 0.5),
  c(7, 10, 0.5),
  c(5, 15, 0.33),
  c(7, 15, 0.33),
  
  c(5, 15, 0.66),
  c(7, 15, 0.66),
  c(3, 5, 0.66),
  c(3, 5, 0.33)
)

N_INTERP_PARAM_BITS <- log2(length(interp_params))

GENE_DEFINITION <- list(
  "NS_ISS_AGES" = issue_age_groups,
  "NS_INTERP" = interp_params,
  "SMK_ISS_AGES" = issue_age_groups,
  "SMK_INTERP" = interp_params
)

N_BITS <- sum(simplify(lapply(GENE_DEFINITION, function(g) { log2(length(g))})))

if (!(as.integer(N_BITS) == N_BITS)) {
  stop("parameter sets must be powers of 2 in length")
}

```

```{r}

bit_str_iter <- as.integer(round(runif(N_BITS,0, 1), 0))

ga_formula <- number_of_deaths ~ -1 +
    preferred_class + 
    gender * 
      splines::ns(attained_age, Boundary.knots = c(25, 95), knots = c(35, 45, 52, 70)) +
    splines::bs(face_amount_band_int, degree = 1, Boundary.knots = c(1, 11), knots = c(2, 7)) + 
  sm_ind +
  ns_ind:AGE_GRPS_NS +
  ns_ind:AGE_GRPS_NS:log_interp_spline_ns +
  sm_ind:AGE_GRPS_SM +
  sm_ind:AGE_GRPS_SM:log_interp_spline_sm 

ga_params <- list()

bit_str_to_mm <- function(bit_str, model_data) {
  
  ga_params <- list()
  
  bit_str_iter <- bit_str
  
  for (parameter_name in names(GENE_DEFINITION)) {
    
    param_msg_len <- log2(length(GENE_DEFINITION[[parameter_name]]))
    
    msg_i <- bit_str_iter[1:param_msg_len]
    msg_idx <- sum(
      msg_i * 2^seq(0, param_msg_len - 1, by=1)
    ) + 1
    
    ga_params[[parameter_name]] <- GENE_DEFINITION[[parameter_name]][msg_idx] %>%
      simplify()
    
    # shift bit string
    bit_str_iter <- bit_str_iter[-(1:param_msg_len)]
  }
  
  if (length(bit_str_iter) != 0) {
    stop("Message error (remaining bits)")
  }
  
  curr_ns_breaks <- ga_params$NS_ISS_AGES
  curr_sm_breaks <- ga_params$SMK_ISS_AGES
  ns_spline_params <- ga_params$NS_INTERP
  sm_spline_params <- ga_params$SMK_INTERP

  ga_df <- model_data %>%
    mutate(
      sm_ind = ifelse(smoker_status == "SMOKER", 1, 0),
      ns_ind = 1 - sm_ind,
      AGE_GRPS_NS = cut(issue_age, breaks = curr_ns_breaks),
      AGE_GRPS_SM = cut(issue_age, breaks = curr_sm_breaks),
      log_interp_spline_ns = interp_spline(duration, 
                                           !!ns_spline_params[1],
                                           !!ns_spline_params[2],
                                           !!ns_spline_params[3]),
      log_interp_spline_sm = interp_spline(duration, 
                                           !!sm_spline_params[1],
                                           !!sm_spline_params[2],
                                           !!sm_spline_params[3]))
  
  ga_mm <- model.matrix(
    ga_formula, data=ga_df
  )
  
  ga_wts <- tibble(
    term_name = colnames(ga_mm)) %>%
    mutate(
      # must be increasing w/ duration -> lower limit == 0
      lower_limits = case_when(
        (stringr::str_detect(term_name, fixed("log_interp_spline"))) &
          (stringr::str_detect(term_name, fixed("AGE_GRPS_"))) ~ 0,
        TRUE ~ -Inf
      ),
      # must be less than 0 at start (issue) -> upper limit == 0
      upper_limits = case_when(
        (stringr::str_detect(term_name, fixed("AGE_GRPS_"))) &
          !(stringr::str_detect(term_name, fixed("log_interp_spline"))) ~ 0,
        TRUE ~ Inf
      )
    )
  
  list(
    "model.matrix" = ga_mm,
    "model.wts" = ga_wts,
    "ga_params" = ga_params
  )
}


```


```{r}

source("0_lib.R")

ga_fitness_2 <- function(bit_string) {
  
  this_mm <- bit_str_to_mm(bit_string, model_data.train)
    
  print("fitting...")
  glmnet_fit <- glmnet::glmnet(
    this_mm$model.matrix,
    model_data.train$number_of_deaths,
    offset = log(model_data.train$expected_deaths_08_vbt_ult),
    family = "poisson",
    intercept = T,
    lower.limits = this_mm$model.wts$lower_limits,
    upper.limits = this_mm$model.wts$upper_limits
  )
  
  aic_bic <- calc_aic_bic_glmnet(glmnet_fit)
  
  fitness_val <- -1 * min(aic_bic$AIC)
  print(sprintf("\tfitness val: %.2f", fitness_val))
  
  fitness_val
}

memo_fn_fitness <- memoise::memoise(ga_fitness_2)

best_model <- GA::ga(type="binary", 
       memo_fn_fitness, 
       popSize=20, 
       nBits = N_BITS, 
       maxiter = 5, 
       parallel = F,
       pmutation = 0.2,
       elitism = 4,
       pcrossover=0.7)

best_model@solution

```

```{r}
final_ga_bit_str <- best_model@solution[1, ]

bit_str_to_mm(final_ga_bit_str, model_data.train)$ga_params

```


### Fit the best model

```{r}

best_mm <- bit_str_to_mm(final_ga_bit_str, model_data.train)

colSums(best_mm$model.matrix)

best_fit <- glmnet::glmnet(
  best_mm$model.matrix,
  model_data.train$number_of_deaths,
  offset = log(model_data.train$expected_deaths_08_vbt_ult),
  family = "poisson",
  intercept = T,
  lower.limits = best_mm$model.wts$lower_limits,
  upper.limits = best_mm$model.wts$upper_limits
)

aic_bic <- calc_aic_bic_glmnet(best_fit)
  
-1 * min(aic_bic$AIC)

```


### Append Predictions

```{r}

BEST_LAMBDA_IDX <- 100
best_lambda_val <- best_fit$lambda[BEST_LAMBDA_IDX]

best_mm.train <- bit_str_to_mm(
  best_model@solution[1, ], 
  model_data.train)

model_data.train[["pred_deaths"]] <- predict(
  best_fit, 
  newx = best_mm.train$model.matrix,
  s = best_lambda_val,
  type="response",
  newoffset = log(model_data.train$expected_deaths_08_vbt_ult)) %>%
  c()

best_mm.test <- bit_str_to_mm(
  best_model@solution[1, ], 
  model_data.test)

model_data.test[["pred_deaths"]] <- predict(
  best_fit, 
  newx = best_mm.test$model.matrix,
  s = best_lambda_val,
  type="response",
  newoffset = log(model_data.test$expected_deaths_08_vbt_ult)) %>%
  c()

list(
  "train A/E" = sum(model_data.train$number_of_deaths) / sum(model_data.train$pred_deaths),
  "test A/E" = sum(model_data.test$number_of_deaths) / sum(model_data.test$pred_deaths)
)


```



### Examine model factors

```{r}

best_coefs <- predict.glmnet(
  best_fit, newx=X_mat, s=best_lambda_val, type = "coefficients"
)

best_coefs
```


```{r}

predictor_groups <- tibble(
  term_name = best_coefs@Dimnames[[1]]
  ) %>%
  mutate(
    term_group = case_when(
      stringr::str_detect(term_name, fixed("Intercept")) ~ "factor_base_rate",
      
      stringr::str_detect(term_name, fixed("preferred_class")) ~ "factor_preferred_class",
      
      stringr::str_detect(term_name, fixed("gender")) |
        stringr::str_detect(term_name, fixed("attained_age")) ~ "factor_gender_x_attained_age",
      
      stringr::str_detect(term_name, fixed("log_interp_spline_ns")) |
        stringr::str_detect(term_name, fixed("log_interp_spline_sm")) ~ "factor_issue_age_x_duration",
      
      stringr::str_detect(term_name, fixed("AGE_GRPS_NS")) |
        stringr::str_detect(term_name, fixed("AGE_GRPS_SM")) ~ "factor_issue_age_x_uwclass",
      
      stringr::str_detect(term_name, fixed("face_amount_band_int")) ~ "factor_face_amount",
      
      stringr::str_detect(term_name, fixed("sm_ind")) |
        stringr::str_detect(term_name, fixed("ns_ind"))  ~ "factor_issue_age_x_uwclass",
    )
  )

predictor_groups %>% view()

```



```{r}

pred_grid <- expand.grid(
  face_amount_band_int = seq(min(df_ph_data.train$face_amount_band_int), max(df_ph_data.train$face_amount_band_int), by=1),
  attained_age = seq(min(df_ph_data.train$attained_age), max(df_ph_data.train$attained_age), by=1),
  duration = seq(min(df_ph_data.train$duration), max(df_ph_data.train$duration), by=1),
  issue_age = seq(min(df_ph_data.train$issue_age), max(df_ph_data.train$issue_age), by=1),
  gender = unique(df_ph_data.train$gender),
  number_of_preferred_classes = unique(df_ph_data.train$number_of_preferred_classes),
  preferred_class = unique(df_ph_data.train$preferred_class),
  smoker_status = unique(df_ph_data.train$smoker_status),
  number_of_deaths = 0, 
  stringsAsFactors = F) %>% 
  filter(
    issue_age + duration - 1 == attained_age,
    ((smoker_status == "NONSMOKER") | ((smoker_status == "SMOKER") & (number_of_preferred_classes == "2"))),
    (
      ((number_of_preferred_classes == "2") & (preferred_class %in% c("1", "2"))) |
      ((number_of_preferred_classes == "3") & (preferred_class %in% c("1", "2", "3")))
    )) 

pred_grid.mm <- bit_str_to_mm(final_ga_bit_str, pred_grid)

X_pred_grid <- cbind(
  rep(1, nrow(pred_grid)),
  pred_grid.mm$model.matrix
)

coeff_diag <- rep(0, best_coefs@Dim[1])
coeff_diag[best_coefs@i + 1] <- best_coefs@x
coeff_diag <- diag(coeff_diag)

pred_mat <- X_pred_grid %*% coeff_diag
colnames(pred_mat) <- best_coefs@Dimnames[[1]]

pred_df <- pred_mat %>%
  as.data.frame() %>%
  mutate(
    pred_idx = row_number()) %>%
  pivot_longer(
    cols=-c(pred_idx), 
    names_to = "term_name", 
    values_to = "term_value") %>%
  inner_join(predictor_groups, by="term_name") %>%
  group_by(pred_idx, term_group) %>%
  summarise(
    total_value = exp(sum(term_value)),
    .groups = "drop"
  ) %>%
  ungroup() %>%
  pivot_wider(
    id_cols = pred_idx, names_from = "term_group", values_from = "total_value"
  )

full_pred_df <- bind_cols(
  pred_grid,
  pred_df
)

```


```{r}

plot_data %>% view()

```


```{r}

plot_data <- full_pred_df %>%
  select(issue_age, smoker_status, duration, factor_issue_age_x_duration, factor_issue_age_x_uwclass) %>%
  distinct() 

#plot_data %>%
#  filter(issue_age == 18) %>% view()

ggplot(plot_data, aes(x=duration, y=factor_issue_age_x_duration * factor_issue_age_x_uwclass, group=issue_age, color=issue_age)) +
  facet_wrap(~ smoker_status) +
  geom_line()

```

```{r}

pred_df

```


