---
title: "Perm Mortality Model"
author: "Mike McPhee Anderson, FSA"
date: "2024-12-18"
output: html_document
---

```{r setup, include=FALSE}

library(glmnet)
library(xgboost)
library(recipes)
library(rpart)
library(rpart.plot)
library(splines)
library(rsample)
library(shapviz)
library(tidyselect)
library(doMC)
library(tidyverse)
library(nloptr)

knitr::opts_chunk$set(echo = TRUE)

```

# Overview

## Models

Within the ILEC data "Perm" (permanent) life insurance category, there
are some major differences in the underwriting / target market that can be inferred via EDA. 
These differences lead us to three distinct models:

 1. *Simplified Issue:* Policies issued after 2005 without any preferred classes are likely simplified
   issue.  These policies are less than $100K face amount, and are likely to have less stringent underwriting.
   
 2. *Fully Underwritten (recent):*  Post-2005 policies are likely to have at least one preferred class, and I recall that blood testing gained traction in the late 1990s.  These policies will also have some underwriting class selection (select period) that is still applicable.
 
 3. *Fully Underwritten (historical):*  Policies prior to 2005 are likely to have little remaining effects from underwriting selection.  Also, this data is unlikely to have preferred classes, and is mostly just smoker vs. non-smoker for underwriting classes.  The majority of it will be smaller face amounts.
 
## Training and Testing Splits

Unfortunately, the ILEC data is aggregated in such a way that makes a typical train / test split cumbersome. While we could sample the exposures and deaths for each aggregated cell, things like fractional exposures and the actuarial custom of setting deaths to have a full year of exposure introduce some opportunities for noise.  

Instead, we'll use the most recent two years of data for testing, and perform cross-validation within the training set by observation year.  This is more closely aligned with what an "unseen observation" is in practice, i.e., another calendar year of data.  

# Setup 

## Load Data

```{r}

face_amount_order <- c(
  "1-9999",
  "10000-24999",
  "25000-49999",
  "50000-99999",
  "100000-249999",
  "250000-499999",
  "500000-999999",
  "1000000-2499999",
  "2500000-4999999",
  "5000000-9999999",
  "10000000+"
)

df_study_data <- read_rds("ilec_perm_recent.rds") %>%
  mutate(
    train_test = ifelse(observation_year >= 2016, "TEST", "TRAIN"),
      # no selection period, just ultimate mortality
    expected_deaths_08_vbt_ult = coalesce(qx * policies_exposed),
    # create ordered factor + integer representation of face amount bands
    face_amount_band = ordered(face_amount_band, levels=!!face_amount_order),
    face_amount_band_int = as.integer(face_amount_band),
    issue_age = attained_age - duration + 1
  ) %>%
  filter(expected_deaths_08_vbt_ult > 0)

```

# Perm Historical Model

## Train Test Split

```{r}

df_ph_data <- df_study_data 

df_ph_data.train <- df_ph_data %>% 
  filter(train_test == "TRAIN")

df_ph_data.test <- df_ph_data %>%
  filter(train_test == "TEST")

df_ph_data %>%
  group_by(train_test) %>%
  summarise(
    n_deaths = sum(number_of_deaths),
    .groups="drop"
  ) %>%
  ungroup() %>%
  mutate(
    pct_deaths = n_deaths / sum(n_deaths)
  )

```



```{r}

summary(df_ph_data.train)

```

```{r}

log_interp_spline <- function(v, interp_start, interp_end, interp_power) {
  c1 <- pmin(1, pmax(0, interp_end - v) / (interp_end - interp_start))^interp_power
  c1 * log(v) + (1-c1) * log(interp_end)
}

curr_ns_breaks <- c(-Inf, seq(45, 75, by=10), Inf)
curr_sm_breaks <- c(-Inf, seq(45, 75, by=10), Inf)
ns_spline_params <- c(10, 15, 0.5)
sm_spline_params <- c(10, 15, 0.5)

group_model_data <- function(df) {
  df %>%
    group_by(attained_age, duration, issue_age, number_of_preferred_classes,
             preferred_class, smoker_status, face_amount_band_int, gender) %>%
    summarise(
      number_of_deaths = sum(number_of_deaths),
      expected_deaths_08_vbt_ult = sum(expected_deaths_08_vbt_ult),
      .groups="drop"
    ) %>%
    ungroup() %>%
    mutate(
      sm_ind = ifelse(smoker_status == "SMOKER", 1, 0),
      ns_ind = ifelse(smoker_status == "NONSMOKER", 1, 0),
      AGE_GRPS_NS = cut(issue_age, breaks = curr_ns_breaks),
      AGE_GRPS_SM = cut(issue_age, breaks = curr_sm_breaks)
    )
}

model_data.train <- df_ph_data.train %>% group_model_data()
model_data.test <- df_ph_data.test %>% group_model_data()

```


```{r}

exp_basis_formula <- number_of_deaths ~ 
  sm_ind * splines::ns(attained_age, 
                              Boundary.knots = c(25, 95), 
                              knots = c(35, 40, 45, 52, 70, 80)) +
  ns_ind:AGE_GRPS_NS +
  sm_ind:AGE_GRPS_SM - 1

default_interp_params <- c(10, 15, 0.5)

log_interp_params <- list(
  "ns_ind:AGE_GRPS_NS(-Inf,45]" = default_interp_params,
  "ns_ind:AGE_GRPS_NS(45,55]" = default_interp_params,                                         
  "ns_ind:AGE_GRPS_NS(55,65]" = default_interp_params,                                              
  "ns_ind:AGE_GRPS_NS(65,75]" = default_interp_params,                                               
  "ns_ind:AGE_GRPS_NS(75, Inf]" = default_interp_params,
  "sm_ind:AGE_GRPS_SM(45,55]" = default_interp_params,
  "sm_ind:AGE_GRPS_SM(55,65]" = default_interp_params,
  "sm_ind:AGE_GRPS_SM(65,75]" = default_interp_params,
  "sm_ind:AGE_GRPS_SM(75, Inf]" = c(5, 10, 6) 
)

create_design_mat <- function(df, exp_basis_formula, log_interp_params) {
  
  mm_train <- model.matrix(
    exp_basis_formula,
    data = model_data.train)
  
  interp_param_mat <- matrix(
    0, 
    ncol = length(log_interp_params),
    nrow = nrow(mm_train))
  
  col_i <- 1
  colnames(interp_param_mat) <- paste0(names(log_interp_params), ":log_interp")
  
  for (col_name in names(log_interp_params)) {
  
    interp_def <- log_interp_params[[col_name]]
    
    print(col_name)
    interp_param_mat[, col_i] <- log_interp_spline(
      model_data.train$duration,
      interp_def[1],
      interp_def[2],
      interp_def[3]
    ) * mm_train[, col_name]
    
    col_i <- col_i + 1
  }

  cbind(mm_train, interp_param_mat)
}

dmat <- create_design_mat(
  model_data.train,
  exp_basis_formula,
  log_interp_params 
) 

dmat %>% colnames()

```



```{r}

intercept_idx <- which(
  stringr::str_detect(colnames(dmat), fixed("AGE_GRPS_")) &
    !(stringr::str_detect(colnames(dmat), fixed("log_interp")))
)

slope_idx <- which(
  stringr::str_detect(colnames(dmat), fixed("AGE_GRPS_")) &
    (stringr::str_detect(colnames(dmat), fixed("log_interp")))
)

model_limits <- tibble(
  upper_limits = rep(Inf, ncol(dmat)),
  lower_limits = rep(-Inf, ncol(dmat)),
  penalty = rep(1, ncol(dmat))
) 

model_limits[intercept_idx, "upper_limits"] <- 0
model_limits[slope_idx, "lower_limits"] <- 0
model_limits[slope_idx, "upper_limits"] <- 1 / log(15)
model_limits[c(intercept_idx, slope_idx), "penalty"] <- 0.1

rownames(model_limits) <- colnames(dmat)

model_fit <- glmnet::glmnet(
  dmat,
  model_data.train$number_of_deaths,
  offset = log(model_data.train$expected_deaths_08_vbt_ult),
  family = "poisson",
  intercept = T,
  lower.limits = model_limits$lower_limits,
  upper.limits = model_limits$upper_limits,
  penalty.factor = model_limits$penalty
)

best_coef <- coef(model_fit)[, 100]

all(best_coef[1 + intercept_idx] < 0) & 
  all(best_coef[1 + slope_idx] > 0)

best_coef %>% view()

```


## Fit Model with L-BFGS

```{r}

make_likelihood <- function(X, y, offset) {
  function(beta) {
    Xbeta <- X %*% beta + offset
    -(sum(y * Xbeta - exp(Xbeta)))
  }  
}

make_gradient <- function(X, y, offset) {
  function(beta) {
    Xbeta <- X %*% beta + offset
    -(crossprod(X, (y - exp(Xbeta))))
  }  
}

exp_basis_formula <- number_of_deaths ~ 
  smoker_status * splines::ns(attained_age, 
                              Boundary.knots = c(25, 95), 
                              knots = c(35, 40, 45, 52, 70, 80)) +
  ns_ind:AGE_GRPS_NS +
  ns_ind:AGE_GRPS_NS:log_interp_spline_ns +
  sm_ind:AGE_GRPS_SM +
  sm_ind:AGE_GRPS_SM:log_interp_spline_sm 

X_mat <- model.matrix(
  exp_basis_formula,
  data=model_data.train
)

y_vec <- model_data.train$number_of_deaths
offset_vec <- log(model_data.train$expected_deaths_08_vbt_ult)

duration_intercept_idx <- which(str_detect(colnames(X_mat), pattern = fixed("AGE_GRPS_")) &
  !(str_detect(colnames(X_mat), pattern = fixed("log_interp_spline"))))

slope_intercept_idx <- which(
  (str_detect(colnames(X_mat), pattern = fixed("log_interp_spline")))
)


```

```{r}

# warmup_X <- X_mat[, -c(duration_intercept_idx, slope_intercept_idx)]
# 
# warmup_fit <- glm.fit(
#   warmup_X,
#   y_vec,
#   offset = offset_vec,
#   family = poisson()
# )
# 
# init_param <- rep(0, ncol(X_mat))
# 
# warmup_idx <- vapply(names(coef(warmup_fit)), 
#        function(x) {
#         which(x == colnames(X_mat))
#       }, c(0))
# 
# 
# init_param[warmup_idx] <- coef(warmup_fit)[warmup_idx]
# init_param[duration_intercept_idx] <- rep(-1, length(duration_intercept_idx))
# init_param[slope_intercept_idx] <- rep(1 / log(15), length(duration_intercept_idx))
#  
# length(init_param)
# 

```

```{r}

# coeff_diag <- rep(0, best_coefs@Dim[1])
# coeff_diag[best_coefs@i + 1] <- best_coefs@x
# coeff_diag <- diag(coeff_diag)

X_warmup <- X_mat[,-c(1)]

warmup_dur_int_idx <- duration_intercept_idx - 1
warmup_slope_int_idx <- slope_intercept_idx - 1

warmup_limits <- tibble(
  upper_limits = rep(Inf, ncol(X_warmup)),
  lower_limits = rep(-Inf, ncol(X_warmup)),
  penalty = rep(1, ncol(X_warmup))
) 

warmup_limits[warmup_dur_int_idx, "upper_limits"] <- 0
warmup_limits[warmup_slope_int_idx, "lower_limits"] <- 0
warmup_limits[warmup_slope_int_idx, "upper_limits"] <- 1 / log(15)
warmup_limits[c(warmup_dur_int_idx, warmup_slope_int_idx), "penalty"] <- 0.2

rownames(warmup_limits) <- colnames(X_warmup)

warmup_fit <- glmnet::glmnet(
  X_warmup,
  y_vec,
  offset = offset_vec,
  family = "poisson",
  intercept = T,
  lower.limits = warmup_limits$lower_limits,
  upper.limits = warmup_limits$upper_limits,
  penalty.factor = warmup_limits$penalty
)

```

```{r}

init_param <- coef(warmup_fit)[,100] 

check_init_params <- bind_cols(
  tibble(param_val = init_param[-1]), warmup_limits)

view(check_init_params)

```


```{r}

length(init_param)

```
```{r}

BEST_IDX <- 100

predict(warmup_fit,
        newx = X_warmup,
        newoffset = offset_vec,
        s = warmup_fit$lambda[BEST_IDX],
        type = "response") %>%
  c() %>%
  sum()

```

```{r}

sum(y_vec) / sum(exp(offset_vec))

```

```{r}

coef(warmup_fit)[, BEST_IDX][duration_intercept_idx]

```


```{r}

coef(warmup_fit)[, BEST_IDX][slope_intercept_idx]

```


```{r}

exp(coef(warmup_fit)[, BEST_IDX][duration_intercept_idx] + 
     coef(warmup_fit)[, BEST_IDX][slope_intercept_idx] * log(1))

```

```{r}

exp(coef(warmup_fit)[, BEST_IDX][duration_intercept_idx] + 
     coef(warmup_fit)[, BEST_IDX][slope_intercept_idx] * log(15))

```

```{r}

sum(model_data.train$number_of_deaths) / sum(exp(X_mat %*% init_param + log(model_data.train$expected_deaths_08_vbt_ult)))

```

```{r}

intercept_adjustment <- -1*(coef(warmup_fit)[, BEST_IDX][duration_intercept_idx] + 
     coef(warmup_fit)[, BEST_IDX][slope_intercept_idx] * log(15))

init_param[duration_intercept_idx] <- init_param[duration_intercept_idx] + intercept_adjustment

```


```{r}

select_param_idx <- c(duration_intercept_idx, slope_intercept_idx)

optim_init_params <- init_param[-select_param_idx]
offset_params <- init_param[select_param_idx]

X_optim <- X_mat[, -select_param_idx]
X_offset <- X_mat[, select_param_idx]

new_offset_vec <- offset_vec + X_offset %*% offset_params

```



```{r}

obj_fn <- make_likelihood(
  X_optim, y_vec, new_offset_vec
)

grad_fn <- make_gradient(
  X_optim, y_vec, new_offset_vec
)

```


```{r}

hin <- function(x) {
  ineq_vals <- rep(0, length(x))
  # ineq_vals[duration_intercept_idx] <- x[duration_intercept_idx]
  # ineq_vals[slope_intercept_idx] <- -x[slope_intercept_idx]
  ineq_vals
}

heq <- function(x) { 
  eq_vals <- rep(0, length(x))
  # eq_vals[25] <- x[17] + log(15)*x[25]
  # eq_vals[26] <- x[18] + log(15)*x[26]
  # eq_vals[27] <- x[19] + log(15)*x[27]
  # eq_vals[28] <- x[20] + log(15)*x[28]
  # eq_vals[29] <- x[21] + log(15)*x[29]
  # eq_vals[30] <- x[22] + log(15)*x[30]
  # eq_vals[31] <- x[23] + log(15)*x[31]
  # eq_vals[32] <- x[24] + log(15)*x[32]
  eq_vals
}

this_init <- optim_init_params
this_init[1] <- 0

optim_res <- nloptr::auglag(this_init, 
       obj_fn, 
       gr = grad_fn, 
       hin = hin, 
       heq = hin, 
       deprecatedBehavior = FALSE,
       control=nl.opts(list(maxeval = 10000)),
       localsolver = "LBFGS")

optim_res

```

```{r}

optim_res$par - optim_init_params

```

```{r}

sum(model_data.train$number_of_deaths) / 
  sum(exp(X_optim %*% optim_res$par + new_offset_vec))

```

```{r}

model_data.train[["final_pred"]] <- exp(X_optim %*% optim_res$par + new_offset_vec)

```


```{r}

plot_data <- model_data.train %>%
  group_by(attained_age, smoker_status) %>%
  summarise(
    ae = sum(number_of_deaths) / sum(final_pred),
    .groups="drop"
  ) %>%
  ungroup()


ggplot(plot_data, aes(x=attained_age, y=ae, color=smoker_status, group=smoker_status)) +
  geom_line()


```
```{r}


plot_data <- model_data.train %>%
  group_by(AGE_GRPS_NS, smoker_status, duration) %>%
  summarise(
    ae = sum(number_of_deaths) / sum(final_pred),
    .groups="drop"
  ) %>%
  ungroup()


ggplot(plot_data, aes(x=duration, y=ae, color=smoker_status, group=smoker_status)) +
  facet_wrap(~ AGE_GRPS_NS, scales = "free_y") +
  geom_line() +
  geom_hline(yintercept = 1) +
  theme_bw()


```

