---
title: "Perm Mortality Model"
author: "Mike McPhee Anderson, FSA"
date: "2024-12-18"
output: html_document
---

```{r setup, include=FALSE}

library(glmnet)
library(xgboost)
library(recipes)
library(rpart)
library(rpart.plot)
library(splines)
library(rsample)
library(shapviz)
library(tidyselect)
library(doMC)
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)

```

# Overview

## Models

Within the ILEC data "Perm" (permanent) life insurance category, there
are some major differences in the underwriting / target market that can be inferred via EDA. 
These differences lead us to three distinct models:

 1. *Simplified Issue:* Policies issued after 2005 without any preferred classes are likely simplified
   issue.  These policies are less than $100K face amount, and are likely to have less stringent underwriting.
   
 2. *Fully Underwritten (recent):*  Post-2005 policies are likely to have at least one preferred class, and I recall that blood testing gained traction in the late 1990s.  These policies will also have some underwriting class selection (select period) that is still applicable.
 
 3. *Fully Underwritten (historical):*  Policies prior to 2005 are likely to have little remaining effects from underwriting selection.  Also, this data is unlikely to have preferred classes, and is mostly just smoker vs. non-smoker for underwriting classes.  The majority of it will be smaller face amounts.
 
## Training and Testing Splits

Unfortunately, the ILEC data is aggregated in such a way that makes a typical train / test split cumbersome. While we could sample the exposures and deaths for each aggregated cell, things like fractional exposures and the actuarial custom of setting deaths to have a full year of exposure introduce some opportunities for noise.  

Instead, we'll use the most recent two years of data for testing, and perform cross-validation within the training set by observation year.  This is more closely aligned with what an "unseen observation" is in practice, i.e., another calendar year of data.  

# Setup 

## Load Data

```{r}

face_amount_order <- c(
  "1-9999",
  "10000-24999",
  "25000-49999",
  "50000-99999",
  "100000-249999",
  "250000-499999",
  "500000-999999",
  "1000000-2499999",
  "2500000-4999999",
  "5000000-9999999",
  "10000000+"
)

df_study_data <- read_rds("ilec_perm_recent.rds") %>%
  mutate(
    train_test = ifelse(observation_year >= 2016, "TEST", "TRAIN"),
      # no selection period, just ultimate mortality
    expected_deaths_08_vbt_ult = coalesce(qx * policies_exposed),
    # create ordered factor + integer representation of face amount bands
    face_amount_band = ordered(face_amount_band, levels=!!face_amount_order),
    face_amount_band_int = as.integer(face_amount_band),
    issue_age = attained_age - duration + 1
  ) %>%
  filter(expected_deaths_08_vbt_ult > 0)

```

# Perm Historical Model

## Train Test Split

```{r}

df_ph_data <- df_study_data 

df_ph_data.train <- df_ph_data %>% 
  filter(train_test == "TRAIN")

df_ph_data.test <- df_ph_data %>%
  filter(train_test == "TEST")

df_ph_data %>%
  group_by(train_test) %>%
  summarise(
    n_deaths = sum(number_of_deaths),
    .groups="drop"
  ) %>%
  ungroup() %>%
  mutate(
    pct_deaths = n_deaths / sum(n_deaths)
  )

```



```{r}

summary(df_ph_data.train)

```

### Knot Selection

```{r}

group_model_data <- function(df) {
  df %>%
    group_by(attained_age, duration, issue_age, number_of_preferred_classes,
             preferred_class, smoker_status, face_amount_band_int, gender) %>%
    summarise(
      number_of_deaths = sum(number_of_deaths),
      expected_deaths_08_vbt_ult = sum(expected_deaths_08_vbt_ult),
      .groups="drop"
    ) %>%
    ungroup()
}

model_data.train <- df_ph_data.train %>% group_model_data()
model_data.test <- df_ph_data.test %>% group_model_data()

```




```{r}

interp_spline <- function(v, interp_start, interp_end, interp_power) {
  c1 <- pmin(1, pmax(0, interp_end - v) / (interp_end - interp_start))^interp_power
  c1 * v + (1-c1) * v
}

prep_model_data <- function(df, curr_ns_breaks, curr_sm_breaks, 
                            ns_spline_params, sm_spline_params) {
  df %>%
    mutate(
      sm_ind = ifelse(smoker_status == "SMOKER", 1, 0),
      ns_ind = 1 - sm_ind,
      AGE_GRPS_NS = cut(issue_age, breaks = curr_ns_breaks),
      AGE_GRPS_SM = cut(issue_age, breaks = curr_sm_breaks),
      log_interp_spline_ns = log(pmin(15, duration)),
      log_interp_spline_sm = log(pmin(15, duration)))
      # log_interp_spline_ns = interp_spline(log(duration), 
      #                                      !!ns_spline_params[1],
      #                                      !!ns_spline_params[2],
      #                                      !!ns_spline_params[3]),
      # log_interp_spline_sm = interp_spline(log(duration), 
      #                                      !!sm_spline_params[1],
      #                                      !!sm_spline_params[2],
      #                                      !!sm_spline_params[3]))

}

age_breaks <- c(-Inf, seq(35, 65, by=10), Inf)
interp_params <- c(10, 15, 0.5)

model_data.train <- model_data.train %>%
  prep_model_data(
    age_breaks, age_breaks,
    interp_params, interp_params)

model_data.test <- model_data.test %>%
  prep_model_data(
    age_breaks, age_breaks,
    interp_params, interp_params)

ga_formula <- number_of_deaths ~
  ns_ind:AGE_GRPS_NS +
  ns_ind:AGE_GRPS_NS:log_interp_spline_ns +
  sm_ind:AGE_GRPS_SM +
  sm_ind:AGE_GRPS_SM:log_interp_spline_sm - 1

ga_mm <- model.matrix(
  ga_formula, data=model_data.train
)

print("fitting...")
glmnet_fit <- glmnet::glmnet(
  ga_mm,
  model_data.train$number_of_deaths,
  offset = log(model_data.train$expected_deaths_08_vbt_ult),
  family = "poisson",
  intercept = T
)

source("0_lib.R")
aic_bic <- calc_aic_bic_glmnet(glmnet_fit)

min(aic_bic$AIC)
```

```{r}

colnames(ga_mm)

```


### Append Predictions

```{r}

BEST_LAMBDA_IDX <- 50
best_lambda_val <- glmnet_fit$lambda[BEST_LAMBDA_IDX]

model_data.train[["pred_deaths"]] <- predict(
  glmnet_fit, 
  newx = model.matrix(ga_formula, model_data.train),
  s = best_lambda_val,
  type="response",
  newoffset = log(model_data.train$expected_deaths_08_vbt_ult)) %>%
  c()

model_data.test[["pred_deaths"]] <- predict(
  glmnet_fit, 
  newx = model.matrix(ga_formula, model_data.test),
  s = best_lambda_val,
  type="response",
  newoffset = log(model_data.test$expected_deaths_08_vbt_ult)) %>%
  c()

list(
  "train A/E" = sum(model_data.train$number_of_deaths) / sum(model_data.train$pred_deaths),
  "test A/E" = sum(model_data.test$number_of_deaths) / sum(model_data.test$pred_deaths)
)


```



### Examine model factors

```{r}
best_coefs <- predict.glmnet(
  glmnet_fit, newx=X_mat, s=best_lambda_val, type = "coefficients"
)

predictor_groups <- tibble(
  term_name = best_coefs@Dimnames[[1]]
  ) %>%
  mutate(
    term_group = "factor_issue_age_x_duration"
  )

predictor_groups %>% view()

```

```{r}
best_coefs
```

```{r}

pred_grid <- expand.grid(
  duration = seq(min(df_ph_data.train$duration), max(df_ph_data.train$duration), by=1),
  issue_age = seq(min(df_ph_data.train$issue_age), max(df_ph_data.train$issue_age), by=1),
  smoker_status = unique(df_ph_data.train$smoker_status),
  number_of_deaths = 0, 
  stringsAsFactors = F) %>% 
  prep_model_data(
    age_breaks, age_breaks,
    interp_params, interp_params
  )


X_pred_grid <- cbind(
  rep(1, nrow(pred_grid)),
  model.matrix(ga_formula, data = pred_grid)
)

coeff_diag <- rep(0, best_coefs@Dim[1])
coeff_diag[best_coefs@i + 1] <- best_coefs@x
coeff_diag <- diag(coeff_diag)

pred_mat <- X_pred_grid %*% coeff_diag
colnames(pred_mat) <- best_coefs@Dimnames[[1]]

pred_df <- pred_mat %>%
  as.data.frame() %>%
  mutate(
    pred_idx = row_number()) %>%
  pivot_longer(
    cols=-c(pred_idx), 
    names_to = "term_name", 
    values_to = "term_value") %>%
  inner_join(predictor_groups, by="term_name") %>%
  group_by(pred_idx, term_group) %>%
  summarise(
    total_value = exp(sum(term_value)),
    .groups = "drop"
  ) %>%
  ungroup() %>%
  pivot_wider(
    id_cols = pred_idx, names_from = "term_group", values_from = "total_value"
  )

full_pred_df <- bind_cols(
  pred_grid,
  pred_df
)

```


```{r}

plot_data <- full_pred_df %>%
  select(issue_age, smoker_status, duration, factor_issue_age_x_duration) %>%
  distinct() %>%
  mutate(factor_issue_age_x_duration = factor_issue_age_x_duration) %>%
  filter(issue_age == 45)

plot_data %>% view()

#plot_data %>%
#  filter(issue_age == 18) %>% view()

ggplot(plot_data, aes(x=duration, y=factor_issue_age_x_duration, group=issue_age, color=issue_age)) +
  facet_wrap(~ smoker_status) +
  geom_line()

```
